A* with Perfect Potentials

### 2020-04-11 paper/ch_potentials.tex
foo

\section{Introduction}
\label{sec:intro}
Route planning in street networks of continental scale is a well-researched topic with many applications~\cite{bdgmpsww-rptn-16}.
It is usually formalized as a variant of the classical shortest path problem in weighted graphs.
Efficiently computing flexible and optimal routes is a thought challenge.
Optimal means that the computed path always has minimum cost.
Being able to change the edge weights is flexibility.
Efficiently means that the path computation is fast.

Dijkstra's algorithm \cite{?} achieves flexiblity and optimality but unfortunately rrequires several seconds to compute paths across continental road networks, which is too slow for many applications.
A classic speedup approach is the well-known A* algorithm \cite{hnr-afbhd-68}.
While flexible, A*'s performance crucially depends on the quality of the potential, also called heuristic, function.
Finding such a function is difficult.
A lot of research of the last decade has therefore focused on mostly hierarchical approaches \cite{CH,CCH,CRP,HIGHWAYHIERACHIES,MLD} which do not require such a function.
A popular hierachical algorithm is Contraction Hierarchy (CH) \cite{?} and its Live-Traffic capable Customizable CH (CCH) \cite{?} extension.
While undoubtedly fast, hierarchical approaches sacrifice flexibility to obtain speed.
In this paper, we explore CH-Potentials, a novel combination of A* and the hierachical techniques CH and CCH.
Our algorithm is built upon ideas from Phast \cite{?}, a CH-extension.
The novel idea is to compute optimal A* potentials lazily.
The hierarchical approach is applied to the lower bound graph with respect to the edge costs.
As the hierarchical approaches are optimal, the resulting A* potential is the tightest possible.
With this approach, we combine some of the hierarchical speed with the flexibility offered by A* - the best of both worlds - while retaining optimality.
Figure~\ref{img:search-space} depicts an example  A* search with CH-Potentials.

@@

We observe, that a potential evaluation is only necessary for nodes that are placed into the queue.
We therefore avoid placing nodes into the queue.

@@

Similarily to the basic A* algorithm, our algorithm continously removes nodes from the queue $Q$.
Denote by $x$ such a node.
Contrary to A*, our algorithm does not only consider the immediate outgoing edges of $x$.
Instead, a queue-less backtracking search at $x$ is started.
For every node $y$ visited in the backtracking search, the following steps are performed:

If $D[y]$ is less than the distance along the path of the backtracking search, the search is pruned.
Otherwise, $D[y]$ is reduced.

If $y$ is already in $Q$, then the backtracking search is pruned and the key of $y$ is decreased, if appropriate.
Otherwise, we do some magic and sometimes continue the search and sometimes prune.

TODO: description incomplete

\section{Largest Biconnected Component}

@@

>>>>

\section{Introduction}
\label{sec:intro}
Route planning in street networks of continental scale is a well-researched topic with many applications~\cite{bdgmpsww-rptn-16}.
It is usually formalized as a variant of the classical shortest path problem in weighted graphs.
Efficiently computing flexible and optimal routes is a thought challenge.
Optimal means that the computed path always has minimum weight.
Efficiently means that the path computation is fast.
Being able to change the edge weights between path computations is flexibility.

Dijkstra's algorithm \cite{?} achieves flexiblity and optimality but unfortunately rrequires several seconds to compute paths across continental road networks, which is too slow for many applications.
A classic speedup approach is the well-known A* algorithm \cite{hnr-afbhd-68}.
While flexible, A*'s performance crucially depends on the quality of the potential, also called heuristic, function.
Finding such a function that can be computed efficiently is difficult.
A lot of research of the last decade has therefore focused on mostly hierarchical approaches \cite{CH,CCH,CRP,HIGHWAYHIERACHIES,MLD} which do not require such a function.
A popular hierachical algorithm is Contraction Hierarchy (CH) \cite{geisberger}.
There is a lot of work that extends CH to handle more complex settings.
For example, in \cite{sabine, kobitzsch} multi-criteria optimization is studied.
In \cite{CCH}, CH is modified to support Live-Traffic.
Historic traffic patterns are combined with CH in \cite{TD-CCH-preprint, TDCH, TDCRP}.
Specialized variants for electric vehicle routing are developped in \cite{moritz baum, charge}.
While these works show that it is possible to extend hierarchical approaches, they also show that it is non-trivial.
Further, in every setting the flexibility available at query time is fairly limited and usually encompasses only one very specific setting.

While undoubtedly fast, hierarchical approaches tend to sacrifice flexibility to obtain speed.
In this paper, we explore CH-Potentials, a novel combination of A* and the hierachical techniques CH and CCH.
We are not the first to combine hierarchical approaches and A* \cite{core-alt, real, charge, bdsssw-chgds-08}.
However, past work mostly uses A* to accelerate hierarchical approaches and sacrifzies A*'s felxibility in the process.
Our primary algorithm design objective is to retain A*'s flexibility.

Our algorithm is built upon ideas from Phast \cite{?}, a CH-extension.
The novel idea is to compute optimal A* potentials lazily.
The hierarchical approach is applied to the lower bound graph with respect to the edge costs.
As the hierarchical approaches are optimal, the resulting A* potential is the tightest possible.
With this approach, we combine some of the hierarchical speed with the flexibility offered by A* - the best of both worlds - while retaining optimality.
Figure~\ref{img:search-space} depicts an example  A* search with CH-Potentials.

@@

We observe, that a potential evaluation is always paired with a queue push operation.
By avoiding push operations we can therefore also avoid potential evaluations.

@@

Similarily to the basic A* algorithm, our algorithm continously removes nodes from the queue $Q$.
Denote by $x$ a node removed with the queue.
Contrary to A*, our algorithm does not only consider the immediate outgoing edges of $x$.
Instead, a limited, queue-less backtracking search at $x$ is started.
The idea is to process local tree-like areas of the graph without queue interaction.
We exploit that backtracking is fast on trees.

The backtracking search continues until either the tentative distance at a node is lower than the length of the path found by the search or a stop node is encountered.
Every node that was already pushed is a stop node.
Nodes with degree 4 or more is a stop node (TODO: @Tim out-degree ?).
Every degree 3 node that is reached via another degree 3 node from the backtracking search start is a stop node.
If the search was stopped because of the tentative distance, then it is aborted.
If the search was stopped because a stop node was encountered, the stop node is added to the queue.
This implies a potential evaluation.
If the node is already in the queue, a decrease key operation is executed.
One can view this modifed A* algorithm as a lazy version of \cite{topocore} where the set of virtually contracted nodes is computed at query time.

\subsection{Largest Biconnected Component}

@@

\section{Experiments}

\begin{itemize}
\item Avoid nothing running times, i.e., what is if $w_q = q_\ell$?
\item Multiply all weights by $\alpha$ for $\alpha$ in 1.1, 1.5, 2, 3
\item Avoid highway
\item Avoid tunnel
\item Avoid highway + tunnel
\item MapBox Live Traffic
\item Ptv-TD traffic? Achtung: Das Fass würde ich nicht aufmachen. Am Ende will ein Reviewer, dass wir avoid highway mit TD kombinieren... das ist NP-schwer
\item Running times if OSM degree 2 nodes are not extracted!
\end{itemize}
